A Two-Infall Star Formation History with Outflows and Radial Migration

### 2025-02-07 src/tex/ms.tex

The infall rate is described by two successive, exponentially declining bursts in time. The first infall component induces the formation of the thick disk, and the second component produces the thin disk. At a given galactocentric radius $R_{\rm gal}$, the un-normalized infall rate is
\begin{equation}
    \label{eq:twoinfall-ifr}
    f_{\rm in}(t|R_{\rm gal}) = e^{-t/\tau_1} + f_{2/1}(R_{\rm gal}) e^{-(t-t_{\rm max})/\tau_2},
\end{equation}
where $\tau_1$ and $\tau_2$ are the first and second infall timescales, respectively, $t_{\rm max}$ is the onset of the second infall and thus the time of maximum gas infall, and $f_{2/1}$ is the ratio of the second infall amplitude to the first. We calculate $f_{2/1}$ for each zone such that the resulting stellar density profile follows a two-component disk, with the surface density ratio of the thick and thin disks given by
\begin{equation}
    f_\Sigma(R) \equiv \frac{\Sigma_1(R)}{\Sigma_2(R)} = f_{\Sigma,0} e^{R(1/R_2 - 1/R_1)}.
\end{equation}
We adopt a thick disk scale radius of $R_1=2.0$ kpc, a thin disk scale radius of $R_2=2.5$ kpc, and a central surface density ratio of $f_{\Sigma,0}=0.27$ \citep{bland-hawthorn_galaxy_2016}.

The parameters of Equation \ref{eq:infall-rate} have important effects on the chemical evolution of the model. Previous studies of the two-infall model have adopted values for the infall parameters of $\tau_1\sim0.1-1$ Gyr, $\tau_2\sim4-10$ Gyr, and $t_{\rm max}\sim4$ Gyr. Figure \ref{fig:twoinfall-parameters} illustrates the effect of varying each of these parameters on gas abundance tracks and stellar abundance distributions in a one-zone model. The first infall timescale $\tau_1$ primarily affects the stellar distribution along the high-$\alpha$ sequence, with negligible effect on the low-$\alpha$ sequence (a). On the other hand, the second infall timescale $\tau_2$ controls the width of the MDF and the [O/Fe] distribution of the low-$\alpha$ sequence, with a shorter $\tau_2$ producing a distribution which is skewed to higher [O/Fe] (b). The time of maximum infall $t_{\rm max}$ (c) strongly affects the overall stellar abundance distribution for values $t_{\rm max}\leq2$ Gyr, but in this case the gas tracks do not produce the characteristic abundance loop. For $t_{\rm max}>2$ Gyr, varying $t_{\rm max}$ results in a minor shift to the mean of the MDF and little change to the [O/Fe] distributions, even though the abundance tracks in [O/Fe]--[Fe/H] space appear very different. We find that $\tau_2$ is the most important parameter to tune in order to reproduce the observed stellar abundance distributions.

@@

@@

\subsection{Stellar Abundance Distributions}

@@

In columns (c) and (d) of Figure \ref{fig:ofe-df}, we attempt to mitigate the issue of the intermediate-$\alpha$ peak for the $y/Z_\odot=2$ yield set. Model (c) switches out our fiducial DTD with a power-law DTD, which decreases the timescale for Fe production (see Figure \ref{fig:dtd}). However, as shown by \citet{dubay_galactic_2024}, models with a power-law DTD struggle to reproduce the high-$\alpha$ sequence for a wide range of star formation histories, and the results are no different for this two-infall model. Finally, in model (d) the metallicity of the infalling gas increases to ${\rm [X/H]}_{\rm CGM}=-0.7$ at late times, which results in very similar [O/Fe] distributions to the $y/Z_\odot=1$ case. We assume that the infalling gas has ${\rm [O/Fe]}=0$ at all times; an alternate run with ${\rm [O/Fe]}=+0.3$ shifted the distribution towards higher [O/Fe], worsening agreement with observations. In summary, pre-enriched gas infall may be necessary for the two-infall model to match the observed [O/Fe] distribution across the disk.

\subsection{Age--Abundance Relations}

\begin{figure*}
    \centering
    \includegraphics{figures/abundance_evolution.pdf}
    \caption{Stellar age--abundance relations produced by select multi-zone models. Each point represents a stellar population drawn from the Solar neighborhood near the midplane ($7\leq R_{\rm gal}\leq 9$ kpc, $0\leq |z| \leq 0.5$ kpc) and is color-coded by its birth radius. For visual clarity, we plot a only random mass-weighted sample of \num{10000} points in each panel. The black curve plots the ISM abundance at $R_{\rm gal}=8$ kpc over time. The red line segments plot the median abundance for APOGEE stars in 2 Gyr-wide age bins, and the shaded regions represent the 16th--84th percentiles in each bin. Each column shows results from a different multi-zone model: {\bf (a)} our fiducial model, with $y/Z_\odot=1$, $\sigma_{\rm RM8}=2.7$ kpc, and pristine gas infall; {\bf (b)} a model with higher yields $y/Z_\odot=2$; {\bf (c)} a model with greater radial migration strength $\sigma_{\rm RM8}=5$ kpc; and {\bf (d)} a model which assumes the infalling gas has metallicity ${\rm [O/H]}_{\rm CGM}={\rm [Fe/H]}_{\rm CGM}=-0.7$. {\bf Key takeaway:} the fiducial model shows two discrepancies with the data: a major dilution event at a lookback time of $\sim9$ Gyr and non-zero abundance evolution at late times. Only model (b) with higher yields can mitigate both of these issues.}
    \label{fig:abundance-evolution}
\end{figure*}

The dilution effect discussed in \ref{sec:yields} is clearly seen in the multi-zone model results. Figure \ref{fig:abundance-evolution} shows stellar age--abundance relations for four model outputs.

@@

\begin{figure*}
    \centering
    \includegraphics{figures/delay_time_distribution.pdf}
    \caption{The effect of the SN Ia delay-time distribution (DTD) on chemical abundance tracks and stellar distributions. {\it Left:} The relative SN Ia rate as a function of time after star formation for each DTD model. The points illustrate the median delay times for each DTD. {\it Right:} Output of one-zone models which assume different SN Ia DTDs with the {\tt yZ2} yield set. The infall parameters are fixed at $\tau_1=1$ Gyr, $\tau_2=10$ Gyr, and $\tau_{\rm on}=3$ Gyr. {\bf Key takeaway:} the extent of the low-$\alpha$ sequence can be reduced by a more prompt SN Ia DTD, but the high-$\alpha$ sequence is also shifted to a more metal-poor regime.}
    \label{fig:dtd}
\end{figure*}

Figure \ref{fig:dtd} illustrates the effect of the SN Ia DTD on the abundance predictions by the two-infall model. The fiducial plateau DTD produces the strongest high-$\alpha$ peak as well as the greatest separation between the two low-$\alpha$ peaks. The exponential DTD produces the most stars near Solar [O/Fe], but the low-$\alpha$ sequence is still distinctly bimodal. Finally, the power-law DTD shrinks the low-$\alpha$ loop such that there is only one low-$\alpha$ peak; however, this DTD produces high-$\alpha$ stars at metallicities much lower than observed, as discussed in \citet{dubay_galactic_2024}.

>>>>

The infall rate is described by two successive, exponentially declining bursts in time. The first infall component induces the formation of the thick disk, and the second component produces the thin disk. At a given galactocentric radius $R_{\rm gal}$, the un-normalized form of the infall rate is
\begin{equation}
    \label{eq:twoinfall-ifr}
    f_{\rm in}(t|R_{\rm gal}) = e^{-t/\tau_1} + f_{2/1}(R_{\rm gal}) e^{-(t-t_{\rm max})/\tau_2},
\end{equation}
where $\tau_1$ and $\tau_2$ are the first and second infall timescales, respectively, $t_{\rm max}$ is the onset of the second infall and thus the time of maximum gas infall, and $f_{2/1}$ is the ratio of the second infall amplitude to the first. We calculate $f_{2/1}$ for each zone such that the resulting stellar density profile follows a two-component disk, with the surface density ratio of the thick and thin disks given by
\begin{equation}
    f_\Sigma(R) \equiv \frac{\Sigma_1(R)}{\Sigma_2(R)} = f_{\Sigma,0} e^{R(1/R_2 - 1/R_1)}.
\end{equation}
We adopt a thick disk scale radius of $R_1=2.0$ kpc, a thin disk scale radius of $R_2=2.5$ kpc, and a central surface density ratio of $f_{\Sigma,0}=0.27$ \citep{bland-hawthorn_galaxy_2016}.

@@

% The parameters of Equation \ref{eq:infall-rate} have important effects on the chemical evolution of the model. 
% Previous studies of the two-infall model have adopted values for the infall parameters of $\tau_1\sim0.1-1$ Gyr, $\tau_2\sim3-10$ Gyr, and $t_{\rm max}\sim3-5$ Gyr. 
Figure \ref{fig:twoinfall-parameters} illustrates the effect of varying the infall parameters on gas abundance tracks and stellar abundance distributions in a one-zone model. We note that all our models are normalized to produce the same thick-to-thin-disk mass ratio. The first infall timescale $\tau_1$, shown in panel (a), primarily affects the stellar distribution along the high-$\alpha$ sequence. Though $\tau_1$ has an apparently large effect on the size of the low-$\alpha$ loop, the effect on the stellar abundance distribution of the low-$\alpha$ sequence is quite small due to the low number of stars formed between $t\sim3-6$ Gyr. We adopt $\tau_1=1$ Gyr for our fiducial value, in line with \citet{spitoni_galactic_2020} but longer than, e.g., \citet{nissen_high-precision_2020} or \citet{spitoni_apogee_2021}, in order to set the peak of the high-$\alpha$ sequence at ${\rm [O/Fe]}\approx+0.3$.

Panel (b) of Figure \ref{fig:twoinfall-parameters} shows that 
% the second infall timescale $\tau_2$ is the most important parameter to tune in order to reproduce the observed stellar abundance distributions. 
the second infall timescale $\tau_2$ controls the width of the MDF and the low-$\alpha$ [O/Fe] distribution. A shorter $\tau_2$ produces a broader [O/Fe] distribution which is skewed to higher [O/Fe], while a longer $\tau_2$ produces both a narrower low-$\alpha$ sequence and a narrower MDF. We note that our maximum value of $\tau_2=30$ Gyr is very close to a constant infall rate, so a further increase in $\tau_2$ has diminishing returns. Between $\tau_2=3-30$ Gyr, the endpoint of the abundance tracks shifts by $\sim0.2$ dex in [Fe/H] and $\sim0.1$ dex in [O/Fe], which could affect the model's ability to reproduce the present-day abundance of the Solar neighborhood. We adopt a fiducial value of $\tau_2=15$ Gyr for the Solar neighborhood in order to minimize the width of the low-$\alpha$ [O/Fe] distribution while still approaching Solar [Fe/H] at late times. This value is in line with the infall timescale recovered by \citet{spitoni_galactic_2020}, and similar to the local star formation timescale adopted by \citet{johnson_stellar_2021}, but significantly longer than the timescales found by \citet{nissen_high-precision_2020} and \citet{spitoni_apogee_2021}. In \todo{Section X}, we explore the effect of varying $\tau_2$ with radius in multi-zone models.

Finally, panel (c) of Figure \ref{fig:twoinfall-parameters} shows that the time of maximum infall $t_{\rm max}$ (c) strongly affects the overall stellar abundance distribution for values $t_{\rm max}\leq2$ Gyr, but in this case the gas tracks do not produce the characteristic abundance loop. For $t_{\rm max}>2$ Gyr, varying $t_{\rm max}$ results in a minor shift to the mean of the MDF and little change to the [O/Fe] distributions, even though the abundance tracks in [O/Fe]--[Fe/H] space appear very different. The value of $t_{\rm max}$ also slightly adjusts the ISM abundance endpoint, as a longer $t_{\rm max}$ means the chemical evolution ``reset'' from the second infall occurs closer to the present day (see discussion in Section \ref{sec:age-abundance}. We adopt a fiducial value of $t_{\rm max}=4.2$ Gyr so that the second infall occurs at a lookback time of 9 Gyr, younger than the median age of the thick disk in the APOKASC-3 catalog of $9.14\pm0.05$ Gyr \citep{pinsonneault_apokasc-3_2024}. This value is generally in line with previous two-infall studies \citep[e.g.,][]{nissen_high-precision_2020,spitoni_galactic_2020,spitoni_apogee_2021}.

@@

\subsection{Stellar Abundance Distributions}
\label{sec:abundance-distributions}

@@

In columns (c) and (d) of Figure \ref{fig:ofe-df}, we attempt to mitigate the issue of the intermediate-$\alpha$ peak for the $y/Z_\odot=2$ yield set. Model (c) switches out our fiducial DTD with a power-law DTD, which decreases the timescale for Fe production (see Figure \ref{fig:dtd}). However, as shown by \citet{dubay_galactic_2024}, models with a power-law DTD struggle to reproduce the high-$\alpha$ sequence for a wide range of star formation histories, and the results are no different for this two-infall model.

Finally, in model (d) the metallicity of the infalling gas increases to ${\rm [X/H]}_{\rm CGM}=-0.7$ at late times. We choose this value because it is the maximum which can plausibly still reproduce the disk abundances; any higher and the infalling gas would be more metal-rich than the most metal-poor thin disk stars. This model results in very similar [O/Fe] distributions to the $y/Z_\odot=1$ case. We assume that the infalling gas has ${\rm [O/Fe]}=0$ at all times; an alternate run with ${\rm [O/Fe]}=+0.3$ shifted the distribution towards higher [O/Fe], worsening agreement with observations. In summary, pre-enriched gas infall may be necessary for the two-infall model to match the observed [O/Fe] distribution across the disk.

\subsection{Age--Abundance Relations}
\label{sec:age-abundance}

\begin{figure*}
    \centering
    \includegraphics{figures/abundance_evolution.pdf}
    \caption{Stellar age--abundance relations produced by select multi-zone models. Each point represents a stellar population drawn from the Solar neighborhood near the midplane ($7\leq R_{\rm gal}\leq 9$ kpc, $0\leq |z| \leq 0.5$ kpc) and is color-coded by its birth radius. For visual clarity, we plot a only random mass-weighted sample of \num{10000} points in each panel. The black curve plots the ISM abundance at $R_{\rm gal}=8$ kpc over time. The red line segments plot the median abundance for APOGEE stars in 2 Gyr-wide age bins, and the shaded regions represent the 16th--84th percentiles in each bin. Age estimates from APOGEE stars come from \citet{leung_variational_2023}. Each column shows results from a different multi-zone model: {\bf (a)} our fiducial model, with $y/Z_\odot=1$, $\sigma_{\rm RM8}=2.7$ kpc, and pristine gas infall; {\bf (b)} a model with higher yields $y/Z_\odot=2$; {\bf (c)} a model with greater radial migration strength $\sigma_{\rm RM8}=5$ kpc; and {\bf (d)} a model which assumes the infalling gas has metallicity ${\rm [O/H]}_{\rm CGM}={\rm [Fe/H]}_{\rm CGM}=-0.7$.}
    \label{fig:abundance-evolution}
\end{figure*}

The dilution effect discussed in \ref{sec:yields} is clearly seen in the multi-zone model results. Figure \ref{fig:abundance-evolution} shows stellar age--abundance relations for four model outputs. Column (a) plots the output of our fiducial model, with $y/Z_\odot=1$, $\sigma_{\rm RM8}=2.7$ kpc, and $Z_{\rm CGM}=0$. The fiducial model shows two major discrepancies with the \citet{leung_variational_2023} age--abundance relation: a major dilution of $\sim0.5$ dex at a lookback time of $\sim9$ Gyr near where the data show a maximum in [O/H], and non-zero abundance evolution at late times where the data show very little abundance evolution. We also note the present-day metallicity is still sub-Solar in the fiducial model, but this can be corrected by decreasing $\eta$. A higher yield set (column b) mitigates both of these issues by shortening the time it takes the ISM metallicity to rebound post-$t_{\rm max}$, producing a much flatter abundance curve at late times. However, a consequence of these higher yields is a poor fit to the stellar [O/Fe] distribution, as discussed in Section \ref{sec:abundance-distributions}.

The observed rise in the median abundance of stars with ages of $\sim4-8$ Gyr is thought to be due to radial migration, as those stars are more likely to have migrated from the dense inner metal-rich regions of the Galaxy during that time. Although our fiducial model does include a prescription for radial migration, the majority of stars in that age range in column (a) have sub-Solar abundances. Column (c) of Figure \ref{fig:abundance-evolution} presents a model with $y/Z_\odot=1$ but with a stronger migration prescription of $\sigma_{\rm RM8}=5$ kpc. Here, the stars which make up the present-day Solar neighborhood are drawn from a wider range of birth $R_{\rm gal}$, producing a broader abundance distribution for any given age. However, even though this prescription is much stronger than the estimates of, e.g., \citet{frankel_measuring_2018}, the model cannot reproduce the observed age--abundance trends in [O/H] and [Fe/H].

Finally, as in Section \ref{sec:abundance-distributions}, we investigate a model where the gas infall is pre-enriched to ${\rm [O/H]}={\rm [Fe/H]}=-0.7$. As shown in column (d) of Figure \ref{fig:abundance-evolution}, this mitigates but does not completely solve the two discrepancies. The dilution effect of the second infall is reduced to the $\sim0.3$-dex level as the gas which replenishes the Galaxy's reservoir is no longer pristine; however, the width of the stellar abundance distribution is also reduced, \todo{because why?}. The late-time gas abundance evolution is similar to the fiducial model, but it ends at slightly super-Solar metallicity --- an effect which can be compensated by a slightly increased value of $\eta$. Overall, no modification to the $y/Z_\odot=1$ model is able to overcome both the dilution and late-time evolution issues.

@@

% \begin{figure*}
%     \centering
%     \includegraphics{figures/delay_time_distribution.pdf}
%     \caption{The effect of the SN Ia delay-time distribution (DTD) on chemical abundance tracks and stellar distributions. {\it Left:} The relative SN Ia rate as a function of time after star formation for each DTD model. The points illustrate the median delay times for each DTD. {\it Right:} Output of one-zone models which assume different SN Ia DTDs with the {\tt yZ2} yield set. The infall parameters are fixed at $\tau_1=1$ Gyr, $\tau_2=10$ Gyr, and $\tau_{\rm on}=3$ Gyr. {\bf Key takeaway:} the extent of the low-$\alpha$ sequence can be reduced by a more prompt SN Ia DTD, but the high-$\alpha$ sequence is also shifted to a more metal-poor regime.}
%     \label{fig:dtd}
% \end{figure*}

% Figure \ref{fig:dtd} illustrates the effect of the SN Ia DTD on the abundance predictions by the two-infall model. The fiducial plateau DTD produces the strongest high-$\alpha$ peak as well as the greatest separation between the two low-$\alpha$ peaks. The exponential DTD produces the most stars near Solar [O/Fe], but the low-$\alpha$ sequence is still distinctly bimodal. Finally, the power-law DTD shrinks the low-$\alpha$ loop such that there is only one low-$\alpha$ peak; however, this DTD produces high-$\alpha$ stars at metallicities much lower than observed, as discussed in \citet{dubay_galactic_2024}.
